---
title: "How to Make a Scatter Plot"
output:
  html_document:
    keep_md: true
    df_print: paged
---

## Comparing Continuous Data

### Numbers are Fun
Continuous data are numbers, typically from taking measurements, and you can find this data type in many fields. In this entry we'll look at how you can **structure** a collection of such data, **visualise** it, and apply a **statistical test** to compare the values. 

This article is for beginners with R, and it doesn’t matter if you’re a student or a professional. You'll need an installation of R (we generally advise using RStudio as well!), and you won't need do download any external data or packages. Hopefully, you can follow along and simply run the code chunks below for yourself - and go on to perform similar analyses on your own data. 

Check our other articles for more R!

# 0. Load Some Built-In Data

* R ships with several built in data sets which we can lean on, and lots of them are great for comparing continuous data.
* We'll be using one called `airquality` - measurements from New York in 1973.
* More details can be found here: https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/airquality.html
* The `data()` function is used to load a built-in dataset.

```{r Load airquality}
data(airquality)
```

There are lots of other datasets available with R, and they're great for practicing various data science tasks. 

# 1. Examine the Data

### The Airquality Dataset

We can take a quick look at this data by applying a couple of handy functions. 

```{r quick look at the data}
head(airquality) # look at the top 5 rows of the data
str(airquality) # how is this data structured
```

We can see that this is a `data.frame` object, with a number of data columns.
The measurements we are interested in are the first four columns, which describe: 

* mean ozone readings 
* solar radiation
* average wind speed
* maximum daily temperature

The final two columns detail when the measurements were collected.

In case you're interested in individual stats for each variable, the `summary()` function enables us to easily explore some summary statistics for these data.
```{r summary airquality}
summary(airquality)
```

Of note here is that we can see how the `Ozone` and `Solar.R` variables in the data have some missing values (NA's), which we will need to pay attention to later on.

# 2. Make a Basic Plot

It's often very helpful to make a quick graph to get a feel for your data. Here we can use the `plot()` function to quickly compare two variables.
```{r fig.height=5, fig.width=5,fig.align='center'}
plot(x = airquality$Wind, # the data for the x-axis
     y = airquality$Temp, # the data for the y-axis
     main = "A Quick Comparison") # the title of the plot
```
From a quick eyeball, it looks like as wind speed increases, temperature decreases. What about the other variables in this dataset?

The `plot()` function is clever, and will make multiple correlations if we ask it to. We can simply give it multiple columns, using the square bracket notation -  `[,1:4]` - and it will plot every column against every other column.
```{r fig.height=6, fig.width=6,fig.align='center'}
plot(airquality[c(1:4)])
```

# 3. Tweak the Plot to Your Liking
```{r fig.height=5, fig.width=5,fig.align='center'}
plot(x = airquality$Wind, # the data for the x-axis
     y = airquality$Temp, # the data for the y-axis
     xlim = c(0,25), # set the x-axis limits
     ylim = c(50,100), # set the y-axis limits
     xlab = "Wind Speed (mph)", # name the x-axis,
     ylab = "Temperature (F)", # name the y-axis,
     pch = 16, # shape of the points 
     col = "slategrey", # colour for the points
     axes = FALSE,
     main = "A Quick Comparison") # the title of the plot
axis(1,seq(0,25,5))
axis(2,seq(50,100,10),las=1)

```


# 4. Do a Statistical Test

These variables can be easily compared to each other with a correlation test, using the `cor()` function.
```{r}
cor(x = airquality$Wind, y = airquality$Temp)
```
By default, this function calculates the 'pearson' *r* value, which measures a linear association between two variables, and ranges between -1 and 1.

* A 1 would be a perfect 'positive' correlation, indicating that as the x-axis variable increases, so does the y-axis variable
* Values closer to 0 indicate a poor or non-existant correlation.
* A value of -1 indicates a perfect 'negative' correlation, with the x-axis variable increasing and the y-axis variable decreasing. 

Values in between, like the `-0.4579879` above, suggest some degree of dispersion from a perfect line. 